### 2018-08-27 paper/preliminaries.tex
updates

First, we introduce the CSP notation and refinement notions. We then discuss
a graph-based semantics for finite-state CSP processes.

@@

\subsubsection*{Communicating Sequential Processes (CSP)} is a process algebra 
supporting system development by refinement. Using CSP, we model both systems and
their components using processes. There are characterised by their patterns
of interactions, modelled by synchronous, instantaneous, and atomic events.

@@

\begin{example}\label{example:CSP}
  We consider the processes $P$, $Q$, and $R$ below. $P$ is initially
  ready to engage in the event $a$, and then makes a choice to
  behave as $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators allow us to define a choice over a
collection of processes $P(i)$.

@@

In fact, in every semantic model, subset containment is used to define refinement. The
model we focus here first is the failures model, which captures both
sequences of interactions and deadlock behaviour. A failure of a process $P$
is a pair $(s,X)$ containing a trace $s$ of $P$ and a refusal:~a set $X$ of
events in which $P$ may refuse to engage, after having performed the events of
$s$. The failures model of a process $P$ records all its failures in a set
$\fails(P)$. 
Using notation $P/s$ to denote the process $P$ after having engaged into trace $s$,
the set 
$\refs(P/s) \defs \{~X | (s,X) \in \fails(P)~\}$ contains the \emph{refusals} 
of $P$ after the
trace $t$. 
Refusals are subset-closed~\cite{Hoare:1985:CSP:3921,Roscoe2010}: if $(s,X)$ is a failure of $P$ and $Y\subseteq X$,
then $(s,Y)\in\fails(P)$ and $Y\in\refs(P/s)$ follows.

Failures refinement, $P \lessdet_F Q$, is defined by set containment 
$\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed, this implies
$(s,\varnothing)\in\fails(P)$ for all traces $s\in\trc(Q)$. As a consequence, failures 
refinement implies trace refinement. Therefore, using the additional conformance relation
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
failures refinement can be expressed by $\lessdet_T$ and $conf$.
\begin{equation}\label{thm:fref-tref-conf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}

%
%
%
%but can
%also be expressed in terms of traces refinement and another conformance
%relation called $conf$ and defined below. The set 
%$\refs(P,t) \defs \{~X | (t,X) \in \fails(P)~\}$ used in this
%definition contains the refusals of $P$ after the
%trace $t$.
%%
%\begin{displaymath}
%  Q\ conf\ P \defs \forall t: \trc(P) \cap \trc(Q) @ \refs(Q,t)
%  \subseteq \refs(P,t)
%\end{displaymath}
%%
%Conformance according to $conf$ requires that no new deadlocks are introduced
%in an implementation. The theorem below, proved in~\cite{CG07}, establishes
%that failures refinement amounts to traces refinement and $conf$.
%%
%\begin{theorem}
%  $(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)$
%  \label{thm:fref-tref-conf}
%\end{theorem}
%%
%Process equivalence in each model is given simply by mutual refinement.

@@

As shown in~\cite{Roscoe:1994:CME:197600}, the failures semantics of 
any finite-state CSP process $P$
can be represented by a \emph{normalised transition graph}
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
$$
with nodes $N$, initial node $\ii n\in N$, and process alphabet $\Sigma$. The
partial \emph{transition function} $t$ maps a node $n$ and an event
$e\in\Sigma$ to its successor node $t(n,e)$. If $(n,e)$ is in the domain of
$t$, then there is a transition, that is, an outgoing edge, from $n$ with label $e$, leading to node $t(n,e)$. Normalisation of
$G(P)$ is reflected by the fact that $t$ is a function. The graph construction
in~\cite{Roscoe:1994:CME:197600} implies that all nodes $n$ in $N$ are reachable
sequences of edges labelled by $e_1\dots e_k$ and connecting states
$\ii n,n_1,\dots,n_{k-1},n$, such that
\[
n_1 = t(\ii n,e_1), \quad n_i = t(n_{i-1},e_i),\ i = 2,\dots,k-1,\quad
n= t(n_{k-1},e_k).
\]
%
By construction, $s\in\Sigma^*$ is a trace of $P$, if and only if 
there is a path through $G(P)$ starting  at $\ii n$ whose edge labels
coincide with $s$. In analogy to $\trc(P)$, we use
notation $\trc(G(P))$ for the set of finite, initialised paths
through $G(P)$, each path represented by its finite sequence of edge labels, and we
note that $\trc(P) = \trc(G(P))$. 
Since $G(P)$ is normalised, there is a unique node
reached by applying the events from $s$ one by one, starting in $\ii n$.
Therefore, $G(P)/s$  is also well defined. By $[n]^0$ we denote the
\emph{initials} of $n$:~the set of events occurring as labels in any outgoing
transitions.
$$
[n]^0 = \{ e\in\Sigma~|~(n,e)\in\dom~t \}
$$
The graph construction from~\cite{Roscoe:1994:CME:197600} guarantees that
$[G(P)/s]^0 = [P/s]^0$ for all traces of $P$.

The total function $r$ maps each node $n$ to a set of non-empty subsets of $\Sigma$. 
The graph construction guarantees that $r(G(P)/s)$ represents the maximal refusals
of $P/s$ for all 
$s\in\trc(P)$. As a consequence,  
\begin{equation}\label{eq:refsAndR}
(s,X)\in\fails(P) \iff s\in\trc(G(P)) \wedge \exists R\in r(G(P)/s): X\subseteq R,
\end{equation}
so $G(P)$ allows us to re-construct the failures semantics of $P$.

@@

Summarising, refinement relations between finite-state CSP processes $P, Q$ can be be
expressed by means of their normalised transition graphs 
$$G(P) = ( N_P, {\ii n}_P, \Sigma, t_P : N_P\times\Sigma \pfun N_P, r_P : N_P \fun \mathbb{P}\mathbb{P}(\Sigma))$$ 
and 
$$G(Q) = ( N_Q, {\ii n}_Q, \Sigma, t_Q : N_Q\times\Sigma \pfun N_Q, r_Q : N_Q \fun \mathbb{P}\mathbb{P}(\Sigma))$$
in the following
way. 
\begin{lemma}
  \label{lemma:tgtrcref}
  \begin{eqnarray}
  P \lessdet_T Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P))
  \label{eq:trcrefa}
  \\
  \label{eq:failconf}
  P \lessdet_F Q & \Leftrightarrow & P \lessdet_T Q \wedge P\ conf\ Q
  \\
  \label{eq:failrefa}
  P\ conf\ Q & \Leftrightarrow & \nonumber
  \forall s\in\trc(G(Q))\cap \trc(G(P)), R_Q\in r_Q(G(Q)/s):  \nonumber
  \\ & & \tabd
  \exists R_P\in r_P(G(P)/s): R_Q\subseteq R_P
  \\
  \label{eq:failrefb}
   & \Leftrightarrow & 
    \forall s\in\trc(G(Q))\cap \trc(G(P)), A_Q\in\minaccs(G(Q)/s):  \nonumber
   \\ & & \tabd
  \exists A_P\in\minaccs(G(P)/s): A_P\subseteq A_Q
 \end{eqnarray}
  \xbox
\end{lemma}
%\begin{lemma}
%  \label{lemma:tgtrcref}
%  \begin{eqnarray}
%  P \lessdet_T Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P))
%  \label{eq:trcrefa}
%  \\
%  \label{eq:failrefa}
%  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
%  \\ & & \forall s\in\trc(G(Q)), R_Q\in r_Q(G(Q)/s):  \nonumber
%  \\ & & \tabd
%  \exists R_P\in r_P(G(P)/s): R_Q\subseteq R_P
%  \\
%  \label{eq:failrefb}
%  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
%  \\ & & \forall s\in\trc(G(Q)), A_Q\in\minaccs(G(Q)/s): \nonumber
%  \\ & & \tabd
%  \exists A_P\in\minaccs(G(P)/s): A_P\subseteq A_Q
%  \end{eqnarray}
%  \xbox
%\end{lemma}
%

@@

%, we have a result similar to
%Theorem~\ref{thm:fref-tref-conf}, but rely on the fact that refusal sets are
%subset closed. In other words, if a process may refuse to engage in all
%events in a set $X$, then it can refuse to engage in all events of subsets of
%$X$. So, in considering $conf$, it is enough to consider maximal refusals.
%Finally, (\ref{eq:failrefb}) captures the same result using acceptances,
%rather than refusals, which as already observed, can be calculated by taking
%the complement of refusals. So, acceptances of $Q$ are required to have a
%corresponding smaller set of acceptances in $P$.

Given a finite-state CSP process $P$ and its normalised transition graph
\[
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
\]
suppose that $V\subseteq\Sigma^*$ is a
prefix-closed set  of sequences of events. By $t(\ii n,V)$ we denote the set
\[
t(\ii n,V) = \{ n\in N~|~\exists s\in V: s\in\trc(P)\wedge G(P)/s = n \}
\]
of nodes in $N$ that are reachable in $G(P)$ by applying traces of $V$. The
lemma below specifies a construction method for such sets $V$ reaching {\it
every} node of $N$.

@@

Since $G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such
that $G(P)/u = G(P)/s_1 = \ol n$. Since $s = s_1.e.s_2$ is a trace of $P$ and
$G(P)/s_1 = \ol n$, then $(\ol n,e)$ is in the domain of $t$. So, $ G(P)/u.e
= G(P)/s_1.e = n$ is a well-defined node of $N$ not contained in $t(\ii
n,V)$. Since $u.e\in V\cup V.\Sigma$, $G(P)$ reaches at least the additional
node $n$ under $V\cup V.\Sigma$. This completes the proof. \xbox
\end{proof}
%\fixme{alcc: Explain the structure of the section? Can the above result be in
%a section, rather than the introduction of the section?}
%\fxnote{jp: I have moved this lemma now to the section on preliminaries, this should make the main section on the finite complete test suite easier to read.}

@@

% =========================================================================

%\subsubsection*{Tool Considerations}
%FDR provides an API~\cite{fdrmanual} that can be used to construct normalised
%transition graphs for CSP processes. The FDR graph nodes are labelled by
%\emph{minimal acceptances} instead of maximal refusals as described above.
%Since such a minimal acceptance set is the complement of a maximal refusal,
%the function $r$ mapping states to their refusals can be implemented by
%taking the complements of all minimal acceptances and then building all their
%subsets. For practical applications, the subset closure is never constructed
%in an explicit way; instead, sets are checked with respect to containment in
%a maximal refusal.
%
%@todo

%The maximal refusals in each process state $P/s$
% are the complements of
%the minimal acceptances of the node $n$
%corresponding to $P/s$. As a consequences, all failures
%of $P$ are represented by some $(s,R)$, where $s$ is an initialised path through the transition graph and $R\subseteq (\Sigma-A)$ for some minimal acceptance $A\in ac(n)$,
%such that $n$ is the node corresponding to $P/s$.

%% =========================================================================
%\subsection{Test Cases and Complete Test Suites}
%\label{sec:cspcompletedef}
%% =========================================================================
%
%@todo

@@

To prove ``$2 \Rightarrow 1$'', assume that that Statement~2 of the lemma holds but 
that  $P\ conf\ Q$ does {\it not} hold. 
According to Lemma~\ref{lemma:tgtrcref}, (\ref{eq:failrefb}), there
exists $s\in\trc(P)\cap \trc(Q)$ such that
\[
\exists A_Q\in\minaccs(G(Q)/s): \forall A_P\in\minaccs(G(P)/s): A_P\not\subseteq A_Q
\qquad (*)
\]
Let $A$ be such an acceptance set $A_Q$ fulfilling (*).
Define
\[
\overline H = \bigcup\{ A_P \setminus A~|~A_P\in\minaccs(G(P)/s) \}.
\]
Since $A_P \setminus A \neq\varnothing$ for all $A_P$ because of (*),
$\overline H$ is a hitting set of $\minaccs(G(P)/s)$ which has an  empty
intersection with $A$. 
Minimising $\overline H$ yields   a minimal hitting set $H\in
\text{minHit}(P/s)$ which is {\it not} a hitting set of $\minaccs(G(Q)/s)$, a
contradiction to Assumption~2. This completes the proof of the lemma. \xbox
\end{proof}
%
This result is used in the sequel in Section~\ref{sec:finitecompletefails}.

%\begin{lemma}
%\label{lemma:hsetb}
%
%\end{lemma}
%\begin{proof}
%asd
%\end{proof}
%

>>>>

%First, we introduce the CSP notation and refinement notions. We then discuss
%a graph-based semantics for finite-state CSP processes.

@@

\subsubsection*{Communicating Sequential Processes (CSP)} is a process algebra 
supporting system development by refinement. Using CSP, we model both systems and
their components using processes. They are characterised by their patterns
of interactions, modelled by synchronous, instantaneous, and atomic events.

@@

\begin{example}\label{example:CSP}
  We consider the processes $P$, $Q$, and $R$ below. $P$ is initially
  ready to engage in the event $a$, and then makes an internal  choice to
  behave like either $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators allow us to define a choice over a
collection of processes $P(i)$.

@@

In fact, in every semantic model, subset containment is used to define refinement. The
model we focus on first is the failures model, which captures both
sequences of interactions and deadlock behaviour. A \emph{failure} of a process $P$
is a pair $(s,X)$ containing a trace $s$ of $P$ and a \emph{refusal}:~a set $X$ of
events in which $P$ may refuse to engage, after having performed the events of
$s$. The failures model of a process $P$ records all its failures in a set
$\fails(P)$. 
Using notation $P/s$ to denote the process $P$ after having engaged into trace $s$,
the set 
$\refs(P/s) \defs \{~X | (s,X) \in \fails(P)~\}$ contains the  refusals
of $P$ after the
trace $s$. 
Refusals are subset-closed~\cite{Hoare:1985:CSP:3921,Roscoe2010}: if $(s,X)$ is a failure of $P$ and $Y\subseteq X$,
then $(s,Y)\in\fails(P)$ and $Y\in\refs(P/s)$ follows.

Failures refinement, $P \lessdet_F Q$, is defined by set containment 
$\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed, this implies
$(s,\varnothing)\in\fails(P)$ for all traces $s\in\trc(Q)$. As a consequence, failures 
refinement implies trace refinement. Therefore, using the auxiliary 
conformance relation
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
failures refinement can be expressed by $\lessdet_T$ and $conf$.
\begin{equation}\label{thm:fref-tref-conf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}

@@

As shown in~\cite{Roscoe:1994:CME:197600}, the failures semantics of 
any finite-state CSP process $P$
can be represented by a \emph{normalised transition graph}
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
$$
with nodes $N$, initial node $\ii n\in N$, and process alphabet $\Sigma$. The
partial \emph{transition function} $t$ maps a node $n$ and an event
$e\in\Sigma$ to its successor node $t(n,e)$. If $(n,e)$ is in the domain of
$t$, then there is a transition, that is, an outgoing edge, from $n$ with label $e$, leading to node $t(n,e)$. Normalisation of
$G(P)$ is reflected by the fact that $t$ is a function. The graph construction
in~\cite{Roscoe:1994:CME:197600} implies that all nodes $n$ in $N$ are reachable by
sequences of edges labelled by $e_1\dots e_k$ and connecting states
$\ii n,n_1,\dots,n_{k-1},n$, such that
\[
n_1 = t(\ii n,e_1), \quad n_i = t(n_{i-1},e_i),\ i = 2,\dots,k-1,\quad
n= t(n_{k-1},e_k).
\]
%
By construction, $s\in\Sigma^*$ is a trace of $P$, if and only if 
there is a path through $G(P)$ starting  at $\ii n$ whose edge labels
coincide with $s$. In analogy to $\trc(P)$, we use
notation $\trc(G(P))$ for the set of finite, initialised paths
through $G(P)$, each path represented by its finite sequence of edge labels, and we
note that $\trc(P) = \trc(G(P))$. 
Since $G(P)$ is normalised, there is a unique node
reached by applying the events from $s$ one by one, starting in $\ii n$.
Therefore, $G(P)/s$  is also well defined. By $[n]^0$ we denote the
\emph{initials} of $n$:~the set of events occurring as labels in any outgoing
transitions.
$$
[n]^0 = \{ e\in\Sigma~|~(n,e)\in\dom~t \}
$$
The graph construction from~\cite{Roscoe:1994:CME:197600} guarantees that
$[G(P)/s]^0 = [P/s]^0$ for all traces of $P$.

The total function $r$ maps each node $n$ to a non-empty set of  (possibly empty) subsets of $\Sigma$. 
The graph construction guarantees that $r(G(P)/s)$ represents the maximal refusals
of $P/s$ for all 
$s\in\trc(P)$. As a consequence,  
\begin{equation}\label{eq:refsAndR}
(s,X)\in\fails(P) \iff s\in\trc(G(P)) \wedge \exists R\in r(G(P)/s): X\subseteq R,
\end{equation}
so $G(P)$ allows us to re-construct the failures semantics of $P$.

@@

Summarising, refinement relations between finite-state CSP processes $P, Q$ can be be
expressed by means of their normalised transition graphs 
$$G(P) = ( N_P, {\ii n}_P, \Sigma, t_P : N_P\times\Sigma \pfun N_P, r_P : N_P \fun \mathbb{P}\mathbb{P}(\Sigma))$$ 
and 
$$G(Q) = ( N_Q, {\ii n}_Q, \Sigma, t_Q : N_Q\times\Sigma \pfun N_Q, r_Q : N_Q \fun \mathbb{P}\mathbb{P}(\Sigma))$$
in the following
way. 
\begin{lemma}
  \label{lemma:tgtrcref}
  \begin{eqnarray}
  P \lessdet_T Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P))
  \label{eq:trcrefa}
  \\
  \label{eq:failconf}
  P \lessdet_F Q & \Leftrightarrow & P \lessdet_T Q \wedge P\ conf\ Q
  \\
  \label{eq:failrefa}
  P\ conf\ Q & \Leftrightarrow & \nonumber
  \forall s\in\trc(G(Q))\cap \trc(G(P)), R_Q\in r_Q(G(Q)/s):  \nonumber
  \\ & & \tabd
  \exists R_P\in r_P(G(P)/s): R_Q\subseteq R_P
  \\
  \label{eq:failrefb}
   & \Leftrightarrow & 
    \forall s\in\trc(G(Q))\cap \trc(G(P)), A_Q\in\minaccs(G(Q)/s):  \nonumber
   \\ & & \tabd
  \exists A_P\in\minaccs(G(P)/s): A_P\subseteq A_Q
 \end{eqnarray}
  \xbox
\end{lemma}

@@

% =========================================================================
\subsubsection*{Reachability Under Sets of Traces}
\label{sec:V}
Given a finite-state CSP process $P$ and its normalised transition graph
\[
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)),
\]
suppose that $V\subseteq\Sigma^*$ is a
prefix-closed set  of sequences of events. By $t(\ii n,V)$ we denote the set
\[
t(\ii n,V) = \{ n\in N~|~\exists s\in V: s\in\trc(P)\wedge G(P)/s = n \}
\]
of nodes in $N$ that are reachable in $G(P)$ by applying traces of $V$. The
lemma below specifies a construction method for such sets $V$ reaching {\it
every} node of $N$.

@@

Since $G(P)$ reaches $G(P)/s_1$ under $V$, there exists a trace $u\in V$ such
that $G(P)/u = G(P)/s_1 = \ol n$. Since $s = s_1.e.s_2$ is a trace of $P$ and
$G(P)/s_1 = \ol n$, then $(\ol n,e)$ is in the domain of $t$. So, $ G(P)/u.e
= G(P)/s_1.e = n$ is a well-defined node of $N$ not contained in $t(\ii
n,V)$. Since $u.e\in V\cup V.\Sigma$, $G(P)$ reaches at least the additional
node $n$ under $V\cup V.\Sigma$. This completes the proof. \xbox
\end{proof}

% =========================================================================
\subsubsection*{Graph Products}
\label{sec:GP}

@@

@@

To prove ``$2 \Rightarrow 1$'', assume that that Statement~2 of the lemma holds but 
that  $P\ conf\ Q$ does {\it not} hold. 
According to Lemma~\ref{lemma:tgtrcref}, (\ref{eq:failrefb}), there
exists $s\in\trc(P)\cap \trc(Q)$ such that
\[
\exists A_Q\in\minaccs(G(Q)/s): \forall A_P\in\minaccs(G(P)/s): A_P\not\subseteq A_Q
\qquad (*)
\]
Let $A$ be such an acceptance set $A_Q$ fulfilling (*).
Define
\[
\overline H = \bigcup\{ A_P \setminus A~|~A_P\in\minaccs(G(P)/s) \}.
\]
Since $A_P \setminus A \neq\varnothing$ for all $A_P$ because of (*),
$\overline H$ is a hitting set of $\minaccs(G(P)/s)$ which has an  empty
intersection with $A$. 
Minimising $\overline H$ yields   a minimal hitting set $H\in
\text{minHit}(P/s)$ which is {\it not} a hitting set of $\minaccs(G(Q)/s)$, a
contradiction to Assumption~2. This completes the proof of the lemma. \xbox
\end{proof}
%
